Random Walks across Social Networks: The Markovian Movement of a Stochastic Soccerball

### 2020-07-26 rmd/draft.Rmd
as handed in

Association football is the world's most popular sport [@IPalacios-Huerta2004] and a multibillion dollar industry [@Deloitte2019]. 
Statistical analysis of the sport was pioneered more than fifty years ago [@CReepBenjamin1968], but has only recently begun to make a more widespread impact [@PCintiaEtAl2015]. Such analysis is of potential interest to coaches and scouts working professionally within the game, journalists and media commentators explaining the game, and gamblers and bookmakers putting money on the game. In this paper I build upon previous work applying Social Network Analysis [@TGrund2012]
and Markov Modelling [@JPena2014] to football data, and introduce an original metric for quantifying the contribution of each player to a team's goal-scoring capability.

@@

In this paper, I develop a rigorous account of a football match as a stochastic social network, the graph of which is generated by the movement of the ball from player to player. I then show how the adjacency matrix of a player, considered as a node in a football match's possession sequence graph, can be used to quantify his contribution to the likelihood of the team scoring, which I call *Markov Expected Goals*. Finally, I discuss the potential and pitfalls of this new metric.

@@

The dataset is provided by WyScout, and is available from FigShare.com [@LPappalardoMassucco2019]. It includes more than three million *event-logs* for every 2017-18 match from the five major European leagues, in JSON format (see Figure 1), with  the details of players, teams, and informative tags (such as the crucial 'Goal') numerically encoded. Each event is associated with an event type (such as 'Shot', 'Pass' or 'Free Kick'), identified with a player, and includes a nested array of the locational *positions* of the event itself and its target.

@@

As part of this preprocessing stage I also labelled each event according first to its *in-play sequence*, and then its *possession sequence*. An in-play sequence ends when the ball goes out of play, as indicated by the subsequent event having the *eventName* 'Free Kick' (a category which includes throw-ins and corners, as well as actual free-kicks given for fouls or offsides); or by the final two events (the shot and the attempted save) being labelled with the *tag* 'Goal'. (Although every goal is followed by a 'free kick' from the centre of the pitch, this never seems to be included in the event logs, presumably because they are created from camera footage which focusses on the goal celebrations and tends to miss the immediate restart of play after a goal).

Intuitively, to identify a possession sequence we should just be able to iterate through the event-logs of a given in-play sequence, and note when the team of the player changes. However, since the event-logs include challenges for the ball even when they are unsuccessful, this would lead to systematically undercounting the length of possession sequences. The solution I adopt is to iterate through the event-logs for a single team within an in-play sequence, paying close attention to the location data: if the location of an event is identical to the target-location of the previous event associated with that same team, then I treat is as part of the same possession sequence, otherwise I treat it as a separate sequence.

@@

We can then visualize the location of the event-log data points in the context of a diagrammatic football pitch, to allow for an immediate and accessible view of the events of any match (see Figures 2 & 3).

@@

With possession sequences that end in a goal, it is generally possible to find video highlights of the events under consideration, and confirm that the data we have does correspond to what actually happened.

<img src="figures/YouTubeClip.gif" width=800>
<figcaption>Figure 5: YouTube Clip showing successful conclusion of possession sequence
</figcaption>

@@

We consider the possession sequences of a football match as *walks* of the football across a social graph in which the nodes are players (considered either individually or collectively), and the connecting edges are defined by the movement of the ball, as it travels from player to player. We can then construct the underlying graph for any particular possession sequence (as in Figure 6), or indeed for the whole game.

<img src="figures/PossessionGraph.png">
<figcaption>Figure 6: Graph of Possession Sequence leading to Goal
</figcaption>

@@

<figcaption>
Figure 7: Heatmap of Adjacency Matrix
</figcaption>

@@

It is increasingly common [@TKnutson2018] to visualize a possession network by positioning each node according to the mean location of the player's touches, and varying the appearance (width, colour or transparency) of the edges to show the frequency of that pass (Figure 8). This swiftly communicates a team's formation, and gives a qualitative sense of a team's strategy.

@@

<figcaption>
Figure 8: Possession Networks for Home and Away Team
</figcaption>

@@

<details open>
<summary>Figure 9</summary>
<figure>
----

@@

<figcaption>
Figure 9: Heatmap of Transition Matrix
</figcaption>

@@

In Figure 9 we show the heatmap of the Transition Matrix derived from the Adjacency Matrix shown in Figure 7. There we showed only the adjacency counts for the starting eleven, but now we include the subsitutes, as well as treating the 'Opposition' collectively as a single node. Most significantly, we now include a 'Goal' as a node. In particular, in the terms of Markov theory, we treat it as an *absorbing state* which loops back on itself with certain probability. This allows us to calculate the expected *hitting time* of a goal, given that any particular player has the ball.

>>>>

Association football is the world's most popular sport [@IPalacios-Huerta2004] and a multibillion dollar industry [@Deloitte2019]. 
Statistical analysis of the sport was pioneered more than fifty years ago [@CReepBenjamin1968], but has only recently begun to make a more widespread impact [@PCintiaEtAl2015]. Such analysis is of potential interest to coaches and scouts working professionally within the game, journalists and media commentators explaining the game, and gamblers and bookmakers putting money on the game. In this paper I build upon previous work applying Social Network Analysis [@TGrund2012] and Markov Modelling [@JPena2014] to football data, and introduce an original method for quantifying the contribution of each player to a team's goal-scoring capability.

@@

In this paper, I develop a rigorous account of a football match as a stochastic social network, the graph of which is generated by the movement of the ball from player to player. I then show how the adjacency matrix of a player, considered as a node in a football match's possession sequence graph, can be used to quantify his contribution to the likelihood of the team scoring, which I call *Markov Expected Goals*. Finally, I discuss the potential and pitfalls of this new indicator.

@@

The dataset is provided by WyScout, and is available from [FigShare.com](https://figshare.com/collections/Soccer_match_event_dataset/4415000) [@LPappalardoEtAl2019a]. It includes more than three million *event-logs* for every 2017-18 match from the five major European leagues, in JSON format (see Figure 1), with  the details of players, teams, and informative tags (such as the crucial 'Goal') numerically encoded. Each event is associated with an event type (such as 'Shot', 'Pass' or 'Free Kick'), identified with a player, and includes a nested array of the locational *positions* of the event itself and its target.

@@

As part of this preprocessing stage I also labelled each event according first to its *in-play sequence*, and then its *possession sequence*.

Intuitively, to identify a possession sequence we should just be able to iterate through the event-logs of a given in-play sequence, and note when the team of the player changes. However, since the event-logs include challenges for the ball even when they are unsuccessful, this would lead to systematically undercounting the length of possession sequences.

The solution I adopt is to iterate through the event-logs for a single team within an in-play sequence, paying close attention to the location data: if the location of an event is identical to the target-location of the previous event associated with that same team, then I treat is as part of the same possession sequence, otherwise I treat it as a separate sequence.

@@

We can then visualize the event-log data as points (Figure 2) or trajectories (Figure 3) in the context of a diagrammatic football pitch, to allow for an immediately accessible overview of any match.

@@

@@

We consider the possession sequences of a football match as *walks* of the football across a social graph in which the nodes are players (considered either individually or collectively), and the connecting edges are defined by the movement of the ball, as it travels from player to player. We can then construct the underlying graph for any particular possession sequence (as in Figure 5), or indeed for the whole game.

<img src="figures/PossessionGraph.png">
<figcaption>Figure 5: Graph of Possession Sequence leading to Goal
</figcaption>

@@

<figcaption>
Figure 6: Heatmap of Adjacency Matrix
</figcaption>

@@

It is increasingly common [@TKnutson2018] to visualize a possession network by positioning each node according to the mean location of the player's touches, and varying the appearance (width, colour or transparency) of the edges to show the frequency of that pass (Figure 7). This swiftly communicates a team's formation, and gives a qualitative sense of a team's strategy.

@@

<figcaption>
Figure 7: Possession Networks for Home and Away Team
</figcaption>

@@

@@

<figcaption>
Figure 8: Heatmap of Transition Matrix
</figcaption>

@@

In Figure 8 we show the heatmap of the Transition Matrix derived from the Adjacency Matrix shown in Figure 6. There we showed only the adjacency counts for the starting eleven, but now we include the subsitutes, as well as treating the 'Opposition' collectively as a single node. Most significantly, we now include a 'Goal' as a node. In particular, in the terms of Markov theory, we treat it as an *absorbing state* which loops back on itself with certain probability. This allows us to calculate the expected *hitting time* of a goal, given that any particular player has the ball.
